Implementation of a Brute Force Attack on A5/1 Cryptographical Algorithm in a GPU-based Volunteering Computing Project

### 2017-02-02 Paper PaCT 2017/paper.tex
+ Section 3 complete

\begin{gather*} X^{19}+X^{18}+X^{17}+X^{14}+1, LFSR1;\\X^{22}+X^{28}+1,
	LFSR2;\\ X^{23}+X^{22}+X^{21}+X^{8}+1, LFSR3.\end{gather*}

@@

The description of the algorithm of determination of unknown 11 bits of LFSR 2
provided above makes obvious the possibility to mount a brute force attack on
the A5/1 generator over the search space of $2^{53}$. The simplicity of the
algorithm provides an opportunity to implement it on a specialized
computational architecture. One such implementation was built with FPGAs by
authors of \cite{COPAC_1}. In the following sections we describe our
implementation of this attack for modern GPUs.

>>>>

\begin{gather*} 
	X^{19}+X^{18}+X^{17}+X^{14}+1, LFSR1;
	\\X^{22}+X^{28}+1, LFSR2;
	\\X^{23}+X^{22}+X^{21}+X^{8}+1, LFSR3.
\end{gather*}

@@

The description of the algorithm of determination of unknown 11 bits of LFSR 2
provided above makes obvious the possibility to mount a brute force attack on
the A5/1 generator over the search space of $2^{53}$. The simplicity of the
algorithm provides an opportunity to implement it on a specialized
computational architecture. One such implementation was built with FPGAs by
authors of \cite{COPAC_1}. In the following sections we describe our
implementation of this attack for a modern GPU.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Implementation of the Anderson's Attack in bit-slicing
technique}\label{sec:bitslc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The effeciency of a brute force attack is defined by two parameters: the speed
of checking of the key candidates and the size of the search space. R.
Anderson's idea described above give us a $$2^{53}$$ search space for the A5/1
generator. Instead of using the 'naive' implementation of A5/1 generator, to
speed up the key candidates checking procedure one can opt to use more
sophisticated alternatives. We evaluated the performance of two different fast
implementations of A5/1 generator in \cite{PAVT_2016}. The first one was based
on an idea of precomputation of the states of the registers LFSR 1-3, and
keeping the states in RAM, to use them as the look-up tables. This approach
demonstrated a considerable speed-up against 'naive' implementation. A similiar
method of precomputation of LFSRs was described in \cite{BSW}. However, in
\cite{PAVT_2016} we found its performance inferior to the second
implementation, that is based on the 'bit-slicing technique'. Next we briefly
describe the idea of this technique, and it's application to Anderson's attack.

Modern general-purpose computational architectures are suboptimal for
implementation of many algorithms. The last operate with bits, while CPUs and
GPUs typically operate with 32-256 bit words. As the result of this
discrepancy, the 'naive' implementation of a cryptographical algorithm won't be
fully using the platform's computational resources. For example, to conduct a
'addition modulo 2' (XOR) operation on 2 boolean arguments, a modern CPU will
position the arguments values into the lowest bits of two 32-bit
general-purpose register (GPR), perform the computation and write the result
into the third 32-bit GPR. In effect, 31 out of 32 bits of GPRs stay idle
during this operation. This problem could be solved by using bitwise operations
that work on GPRs like on boolean vectors - digit by digit. A cryptographical
algorithm could be oftentimes represented as a logical scheme built with the
basic logical gates. Coupled with bitwise operations, this representation
allows the computational platform to apply the algorithm to as many sets of
data, as there are bits in it's GPR (e.g. the platform's register capacity).
This method is called 'SIMD within a register' or the 'bit-slicing' technique.
The first mention of bit-slicing technique applied to cryptographical problem
belongs, seemingly, to E.Biham \cite{BIT_SL}.

Now we describe the main idea of the bit-slicing technique. Consider an
arbitrary total boolean function

\begin{gather*} 
	f:\{0,1\}^n\rightarrow\{0,1\}
\end{gather*}

This function can be represented in the form of the Boolean circuit $C(f)$ over
some complete basis $B$. A common example of such basis is
$B=\{\land,\lor,\neg\}$, but we'll use the basis $B=\{\land,\lor,\neg,\oplus\}$
instead, since it better fits our goals.

Now consider the problem of calculating the arbitrary total boolean function
$f:\{0,1\}^n\rightarrow\{0,1\}$ over all $2^n$ possible inputs. For each input
$X\in\{0,1\}^n$ one can calculate the value of $f$ as a superposition of the
basis functions, according to the scheme $C(f)$. We can now select at least one
fixed order of calculation of basis functions from $C(f)$, that results in
getting the value of $f$. Let $m$ be the number of internal nodes in $C(f)$.
Assuming that the calculation of one basis function takes one processor
instruction, the computation of $f$ over all inputs from $\{0,1\}^n$ will take
$m\cdot2^n$ instructions.

A computational architecture that, in a single instruction execution,
calculates many copies of the same function over many different memory cells,
is called the 'single instruction, multiple data' (SIMD) architecture. When a
modern computational device executes a bitwise logical instruction over its
GPRs, it effectively acts as a SIMD device, individual bits of GPRs playing the
role of the individual memory cells of a SIMD device. The calculation order of
functions in the scheme $C(f)$ always stays the same. This makes it possible to
compute this functions over as many inputs, as there are bits in the device's
GPR.  In effect, if $D$ is the device's GPR capacity, we can simultaneously
walk $D$ instances of the scheme $C(f)$, effectively calculating the value of
$f$ for inputs $X_1,...,X_D$.

Let's examine the arbitrary basis function $g$ with arity 2, and the
corresponding internal node of the scheme $C(f)$. We denote it as $G(x_1,x_2)$,
meaning that it has a single output and two inputs, which values are determined
by boolean variables $x_1,x_2$. Next, collate $g$ with three GPRs denoted
$R_1(g),R_2(g),R_3(g)$, each of which is comprised of $D$ single-bit memory
cells, filled in the following way: 
\begin{itemize} 
\item Register $R_1$ contains $D$ values of the variable $x_1$, corresponding
 to $X_1,...,X_D$; 
\item Register $R_2$ contains $D$ values of the variable $x_2$, corresponding
 to $X_1,...,X_D$;
\item Register $R_3$ contains $D$ values of the function $g$, corresponding to
 matching values of $X_1,...,X_D$.
\end{itemize}
Suppose that all $D$ instances of values of $g$ (matching the corresponding
values of $x_1$ and $x_2$) can be computed as a result of a single bitwise
instruction applied to register $R_1(g),R_2(g)$, while their result is put into
register $R_3(g)$.  If this fact is holding true for every basis function in
the scheme $C(F)$, the computation of $f$ for every input from $\{0,1\}^n$ will
require $m\cdot\frac{2^n}{D}$ instructions. This is the key idea of the
bit-slicing technique \cite{fig:bitsl}.

\begin{figure} \includegraphics[width=\linewidth]{./bitslice.png}
	\caption{Simultaneous computation of $D$ instances of scheme $C(f)$ with
	bit-slicing technique} \label{fig:a51gen} 
\end{figure}

We will call the process of computation of the function $f$ on a single input
from $\{0,1\}^n$ a "thread", by analogy with the computational threads in a
SIMD device. So, with the use of bit-slicing technique, it takes $m$
instructions to complete the computation of $D$ threads.

Now we describe the details of bit-slicing implementation of the A5/1
generator. Suppose that a computational device is able to calculate $D$
instances of any function from the basis $B=\{\land,\lor,\neg,\oplus\}$. Each
of $n\in\{1,...,64\}$ cells of LFSR 1-3 gets a matching word $W_n\in\{0,1\}^D$:

\begin{gather*} 
	LFSR1:W_1,...,W_{19}; 
	\\LFSR2:W_{20},...,W_{42}; 
	\\LFSR3:W_{43},...,W_{64}.
\end{gather*}

In a bit-slicing technique, the shift of the LFSR register (LFSR 1 in this
example) will take the following form: 
\begin{gather*} 
	W`_1=W_{19} \oplus W_{18} \oplus W_{17} \oplus W_{14}, 
	\\W`=W_{n-1}, n \in \{2,...,19\},
\end{gather*}
where $\oplus$ is the bitwise addition modulo 2 of boolean vectors of length
$D$ operator. The calculation of the keystream bit will look like:
\begin{gather*} 
	W_{out}=W_{19} \oplus W_{41} \oplus W_{64}.
\end{gather*}

The conditional clocking is somewhat more complex to implement in a bit-slicing
technique. First, to know if the LFSRs should be shifted or not, one needs to
calculate the corresponding clocking flags $F_1,F_2,F_3$ using the majority
function: 
\begin{gather*} 
	W_{maj}=maj(W_8,W_{30},W_{52})=(W_9 \wedge W_{30}) \vee (W_9 \wedge W_{52})
	\vee (W_{30} \vee W_{52}), \\F_1 = W_9 \oplus {\neg W_{maj}}, \\F_2 =
	W_{30} \oplus {\neg W_{maj}}, \\F_3 = W_{52} \oplus {\neg W_{maj}},
\end{gather*} 
(all operations are bitwise operations on the vector of the length $D$).

To implement the conditional clocking of a LFSR one can use the bitwise
counterpart of the "bitselect" function of arity 3: 
\begin{gather*} a,b,c \in \{0,1\}; \\ BS(a,b,c) = \left\lbrace
	\begin{array}{ll} b,a=1, \\ c,a=0.
	\end{array} \right.
\end{gather*}

If the computational architecture lacks the hardware implementation of this
function, it can be emulated with the usage of the standard bitwise functions
corresponding to the functions of the basis $B$ (example for LFSR 1):
\begin{gather*} BS(a,b,c) = (a \wedge b) \vee (\neg a \wedge c).
\end{gather*}

Clocking the LFSR 1 with the use of $BS(a,b,c)$ and corresponding flag $F_1$
looks the following way:
\begin{gather*} 
	W'_1 = BS(F_1, (W_{19} \oplus W_{18} \oplus W_{17} \oplus W_{14}), W_1);
	\\W'_n = BS( F_1, W_{n-1}, W_n), n \in \{2,..,19\}.
\end{gather*}

Some important details about the bit-slicing implementation of the Anderson's
attack should be covered still. The Anderson's attack follows 2 stages:
\begin{enumerate} 
	\item Calculating the values of 11 bits of LFSR 2 lying left of the
	 clocking bit with the information from the remaining 53 bits (guessed) and
		the keystream (known).
	\item Clocking the generator as normal to check if the guessed filling of
	 the generator matches the known keystream.
\end{enumerate}

The irregular clocking of the A5/1 generator makes it generally impossible to
predict how many clockings of the whole generator (bits of keystream) would be
needed to shift LFSR 2 11 times to complete the Stage 1 of the attack.
Therefore, we again put to use the bitselect function to implement the split of
the attack into 2 stages. Since each individual thread should be able to stay
in a stage and advance to the next stage independent of other threads, we
introduce the special boolean vector $\phi = (\phi_1,...,\phi_D)$, called the
"attack stage flag". The thread with number $i, i \in \{1,...,D\}$ being in the
Stage 1 of the attack corresponds to $\phi_i = 0$, and the Stage 2 of the
attack corresponds to $\phi_i=1$. Let $y_1,...,y_{64}$ be the bits of the
keystream analyzed. Now the clocking of the LFSR 2 considers the stage of the
attack through the usage of the attack stage flag: \begin{gather*} W^\ast_{41}
	= BS(\phi, W_{41},(y \oplus W_{19} \oplus W_{64}));\\W'_{20} =
	BS(F_2,(W^\ast_{41} \oplus W_{40}), W_{20});\\W'_n = BS(F_2,W_{n-1},W_n), n
	\in \{20,...,41\}. 
\end{gather*}

Here $W^\ast_{41}$ is a helper vector holding temporary data. $y$ is the
current bit of the keystream, in the form of a vector consisting of $D$ copies
of $k$-th bit of the keystream. In effect, Stage 1 of the attack's goal is to
calculate the last bit of the LFSR 2 from known keystream bits and known
(guessed) last bits of the LFSRs 1 and 3. At the Stage 2 the last bit of the
LFSR 2 is known and the whole generator scheme is clocked as normal. For the
$i$-th thread the attack stage flag $\phi_i$ is set to 1 after the thread's
instance of LFSR 2 was shifted 11 times. To count the number of LFSR 2 shifts
on a per-thread basis, the bit-slicing implementation of an incremental counter
is used.

The Anderson's attack algorithm described above was realized on an NVIDIA GPU
with the use of CUDA SDK 8.0 \cite{CUDA}. The comparison of the performance of
a GPU to a CPU in execution of bit-slicing and LFSR precomputation-based
\cite{BSW} implementations of Anderson's attack is shown in
\cite{tab:gpuspeed}.

\begin{table} \caption{(LOP3.LUT is a special instruction that implements
	arbitrary bitwise arity 3 functions in hardware. We use it as a substitute
	for the "bitselect" instruction.)} 
	\label{tab:gpuspeed} Performance of the different implementations of the
	Anderson's attack (search space is $2^{53}$) on a CPU and a GPU, measured
	in millions of key checks per second.  
	\begin {center} \begin{tabular} {| l | l | l | l |} \hline Computational
		device & Bit-slicing & LFSR precomputation \cite{BSW} \\ \hline CPU
		Intel Core I7 930 & 37 & 7 \\ \hline GPU NVIDIA GTX 1050 Ti & 9180 &
		483 \\ \hline GPU NVIDIA GTX 1050 Ti (LOP3.LUT) & 11950 & - \\ \hline

	\end{tabular}
\end {center}
\end{table}

Data provided in the \Cref{tab:gpuspeed} tells us that even one mid-range
consumer GPU is enough to make the Anderson's attack run time practical (it
will take around 250 hours). A modern computational cluster outfitted with GPUs
will complete the attack in mere minutes or less. The Anderson's attack's
advantage over the rainbow-tables attack \cite{RAINBOW} is the former's ability
to restore the secret key from 64 bits of keystream with the 100\% probability.
It's advantge over the attack described in \cite{COPAC_1} is in a usage of a
consumer-grade off-the-shelf hardware.
